---
title: "Yeast subphylum transcriptomic study design"
author: "Arnaud N'Guessan"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

## 1.Using power analyses to determine the number of species or genes required
### 1A. Comparing the expression variance in the accessory and core genomes on long time scales of evolution
To perform this task, I will compare the variance in expression levels in the accessory genome to the one in the core genome. To account for outliers like rare genes (0-inflated), I will test the differences in the median variance with a Wilcoxon rank sum test. The power of the test will depend on the number of genes. The function sim.ssize.wilcox.test() from the package MKpower can estimate the power of this test. Assuming a small difference (worst case in term of power) and a gamma distribution for the expression variance, sim.ssize.wilcox.test() can simulate the power at different sample sizes. The sample size is the total number of genes (core and accessory) which ranges from ~4,000 to 20,000 (Figure S1A in DOI: 10.1126/science.adj4503). The expression level would be measured by log2(TPM) like in the budding yeast pan-transcriptomic paper (Figure 2C in DOI 10.1038/s41588-024-01769-9). 

```{r}
set.seed(123)
## two-sample, effect size and gamma parameters are the same than within-species (Figure 2C in DOI 10.1038/s41588-024-01769-9) and the number of genes can range from ~4,000 to 20,000
gene_set1 <- function(n) rgamma(n, scale = 0.35, shape = 1)
gene_set2 <- function(n) rgamma(n, scale = 0.6, shape = 1)
print(summary(rgamma(4000/2, scale = 0.35, shape = 1)))
print(summary(rgamma(4000/2, scale = 0.6, shape = 1)))
pwr_wilcox_small_effect_equal_variance <- MKpower::sim.ssize.wilcox.test(rx = gene_set1, ry = gene_set2, n.min=4000/2,n.max = 20000/2, step.size = 600, iter = 5000,type = "two.sample") #n is the sample size in each group
print(pwr_wilcox_small_effect_equal_variance)

```
The power is already maximized at 4,000 genes in total (2000*2 groups) but what if the effect is even smaller.

```{r}

## two-sample, very small effect
gene_set3 <- function(n) rgamma(n, scale = 0.35, shape = 1)
gene_set4 <- function(n) rgamma(n, scale = 0.4, shape = 1)
print(summary(rgamma(4000/2, scale = 0.35, shape = 1)))
print(summary(rgamma(4000/2, scale = 0.4, shape = 1)))
pwr_wilcox_very_small_effect_high_variance_equal <- MKpower::sim.ssize.wilcox.test(rx = gene_set3, ry = gene_set4, n.min=4000/2,n.max = 20000/2, step.size = 600, iter = 5000,type = "two.sample")
print(pwr_wilcox_very_small_effect_high_variance_equal)
```
### 1B. To highlight genes for which expression changes are selected for
For each gene, I will perform a linear regression between the expression variance and selection metrics like dN/dS or growth rate variance.The variance can be measured on species sets. These metrics will be measured on different species sets of constant size. The optimal set size is a trade-off between the ability to capture expression variation and the resources/time required to collect many species. In the linear regression, each point will be a different set so let's find the minimum number of sets required to reach a power >=0.8 if the effect is small. 

```{r}
library(pwr)
#find the power of different number of sets 
#In pwr.f2.test(u =1, v=198, f2=0.14^2, sig.level = 0.05), an f2 of 0.14 represents a small effect, u = 1 predictor and v = 200 samples - 1predictor - 1  (df)
v_tested_set_sizes <- seq(10,1000,10)
v_pwr_linear_regression_at_different_nb_sets <- rep(NA,length(v_tested_set_sizes))
names(v_pwr_linear_regression_at_different_nb_sets) <- as.character(v_tested_set_sizes)
for (current_nb_sets in v_tested_set_sizes){
  v_pwr_linear_regression_at_different_nb_sets[as.character(current_nb_sets)] <- pwr.f2.test(u = 1, v=current_nb_sets-2, f2=0.14^2, sig.level = 0.05)$power
}
#find the minimum number of sets (to reach a power of 0.8)
#assertthat::assert_that(any(v_pwr_linear_regression_at_different_nb_sets>=0.8))
min_nb_sets <- as.integer(names(v_pwr_linear_regression_at_different_nb_sets[v_pwr_linear_regression_at_different_nb_sets>=0.8])[1])
print(min_nb_sets)
#Create a function that finds the minimum number of species to form the number of sets of size k_species_per_set
find_min_nb_species_for_min_nb_sets_of_size_n <- function(the_min_nb_sets,k_species_per_set){
  n_species <- k_species_per_set
  nb_sets <- choose(n = n_species,k = k_species_per_set)
  while (T){
    nb_sets <- choose(n = n_species,k = k_species_per_set)
    if (nb_sets >= the_min_nb_sets){
      break
    }
    n_species <- n_species + 1
  }
  return(n_species)
}
#Find the number of species required to get the optimal number of sets
v_lst_set_sizes <- 2:20
df_min_nb_species_required_at_different_set_size <- NULL# data.frame(total_nb_species=NULL,set_size=NULL,min_number_sets=NULL,stringsAsFactors = F)
for (current_set_size in v_lst_set_sizes){
  df_min_nb_species_required_at_different_set_size <- rbind(df_min_nb_species_required_at_different_set_size,data.frame(set_size=current_set_size,min_number_sets_required=min_nb_sets,min_nb_species_required=find_min_nb_species_for_min_nb_sets_of_size_n(the_min_nb_sets=min_nb_sets,k_species_per_set=current_set_size),nb_sets_obtained_with_min_nb_species=choose(n =find_min_nb_species_for_min_nb_sets_of_size_n(the_min_nb_sets=min_nb_sets,k_species_per_set=current_set_size) ,k=current_set_size),stringsAsFactors = F))
}

print(df_min_nb_species_required_at_different_set_size)
```
With 30 species in the pilot project (24 species from NRRL + 6 we already have), we can use pairwise comparisons (or higher-order combination) to detect correlation between each gene transcriptional variance and selection metrics (gene pairwise dN/dS or the species differences/variance in growth rate/fitness).  dN/dS can be measured for each gene alignment using phylogenetics tools like Hyphy (https://stevenweaver.github.io/hyphy-site/methods/selection-methods/).

## 2. Selecting the species for the transcriptomic analyses
The criteria are:
  * at least  30 species (24 from NRRL+ 6 species that we already have)
  * The species grow FAST in MOST of the environments (convenient for experiments)
  * The species does not form filaments or aggregates (convenient for RNA extraction and growth measurements): Saccharomycotina
  * TThe species span the whole tree (at least 1 species in most of the orders)

```{r,warning=F}
#species that we already have
v_lst_species_we_have <- c("Saccharomyces cerevisiae", "Saccharomyces paradoxus", "Candida glabrata", "Vanderwaltozyma polysporus", "Zygosaccharomyces rouxii", "Kluyveromyces lactis", "Lachancea thermotolerans", "Lachancea kluyveri", "Candida albicans", "Debaryomyces hansenii", "Komagataella phaffi", "Blastobotrys adenivorans", "Metschnikowia hibisci", "Kluyveromyces marxianus", "Saccharomyces pastorianus", "Yarrowia lipolytica", "Saccharomycopsis fermentans", "Saccharomycopsis javanensis", "Saccharomycopsis fibuligera", "Saccharomycopsis fodiens")

#pathogenic species
v_pathogenic_yeast_species <- c("Pichia kudriavzevii","Meyerozyma guilliermondii","Candida parapsilosis","Candida orthopsilosis","Candida metapsilosis","Candida dubliniensis","Candida albicans","Candida tropicalis","Candida auris","Clavispora lusitaniae","Nakaseomyces glabratus")

#species with quantitative growth data
df_growth_Rates_True_yeast <- readxl::read_excel("GrowthRates_Yeasts.xlsx", range = "B1:AA854")
threshold_proportion_defining_most_envs <- 0.6
mtx_growth_Rates_True_yeast <- as.matrix(df_growth_Rates_True_yeast[,c(-1)])
rownames(mtx_growth_Rates_True_yeast) <- df_growth_Rates_True_yeast$Species
mtx_growth_Rates_True_yeast <- t(apply(X = mtx_growth_Rates_True_yeast,MARGIN = 1,FUN = as.numeric))
v_lst_species_with_quantitative_growth_data <- df_growth_Rates_True_yeast$Species
median_growth_rates_across_envs <- apply(X = mtx_growth_Rates_True_yeast,MARGIN = 2,FUN = function(x) quantile(x,0.5,na.rm=T))
v_is_species_growing_and_top50_fastest_in_most_env <- apply(X = mtx_growth_Rates_True_yeast,MARGIN = 1,FUN = function(the_rates) ((sum(the_rates>0,na.rm=T)/length(the_rates))>=threshold_proportion_defining_most_envs)&((sum(the_rates>median_growth_rates_across_envs,na.rm=T)/ncol(mtx_growth_Rates_True_yeast)) >=threshold_proportion_defining_most_envs))
v_lst_species_growing_top50_fastest_in_most_env <- v_lst_species_with_quantitative_growth_data[v_is_species_growing_and_top50_fastest_in_most_env]
#print(sort(v_lst_species_growing_top50_fastest_in_most_env))
#print(vapply(X = sort(intersect(v_lst_species_with_quantitative_growth_data,v_lst_species_we_have)),FUN = function(the_sp) sum(mtx_growth_Rates_True_yeast[the_sp,] > median_growth_rates_across_envs,na.rm = T),FUN.VALUE = 0))
v_candidate_species <- c("Debaryomyces hansenii","Saccharomycopsis fermentans","Saccharomyces paradoxus","Metschnikowia hibisci","Kluyveromyces marxianus","Lachancea kluyveri","Candida blattariae","Wickerhamomyces strasburgensis","Hanseniaspora nectarophila", "Hanseniaspora opuntiae","Cyberlindnera japonica","Cyberlindnera mrakii","Schwanniomyces capriottii","Wickerhamomyces anomalus","Saccharomycopsis amapae","Candida dosseyi","Candida ulmi","Kodamaea ohmeri","Metschnikowia leonuri","Meyerozyma caribbica", "Ambrosiozyma oregonensis","Saturnispora besseyi","Nakazawaea wyomingensis","Nakazawaea peltata","Schwanniomyces polymorphus var. polymorphus","Zygoascus tannicolus","Blastobotrys peoriensis","Sugiyamaella smithiae","Lipomyces kononenkoae","Dipodascus geniculatus")
#Make sure all the species are in the NRRL 
df_NRRL_catalog <- readxl::read_excel("fungi_in_catalog.xlsx")
df_NRRL_catalog_filtered <- subset(df_NRRL_catalog,`RISK GROUP`== 1 & `TYPE STRAIN` == "YES" & `ANIMAL PERMIT` == "NO" & `PLANT PERMIT` == "NO")
remove(df_NRRL_catalog)
assertthat::assert_that(all(v_candidate_species%in%df_NRRL_catalog_filtered$`CURRENT NAME`))
```
```{r,warning=F}
v_candidate_species_that_we_have <- v_candidate_species[1:6]
v_candidate_species_from_NRRL <- v_candidate_species[7:30]

get_NRRL_id <- get_NRRL_id <- function(the_sp) as.character(subset(df_NRRL_catalog_filtered,`CURRENT NAME` == the_sp)$`NRRL NUMBER`[1]) #function(the_sp) paste0(as.character(subset(df_NRRL_catalog_filtered,`CURRENT NAME` == the_sp)$`NRRL NUMBER`),collapse="/") 
names(v_candidate_species_from_NRRL) <- unname(vapply(X = v_candidate_species_from_NRRL,FUN = get_NRRL_id,FUN.VALUE = ""))

print(v_candidate_species_that_we_have)

print(v_candidate_species_from_NRRL)
```

Most of these species grow in more than 10 environment at a high growth rate (>median). 
A collection of diverse environments from which we can choose from:
Allantoin (27 species growing out of the 30 selected), Fructose (28/30), Glucose (30/30), Mannose (29/30), Urea (20/30; chosen because it is not correlated with other traits) and Sucrose (25/30).